---
title: "E04 - Affymetrix microarrays (Analysis of Gene Expression 2020 @ VSCHT)"
author: "Jiri Novotny"
institute: "Laboratory of Genomics and Bioinformatics @ Institute of Molecular Genetics of the ASCR"
output:
  html_document:
    # code_folding: "show"
    df_print: "paged"
    toc: true
    toc_float: true
    toc_collapsed: false
    toc_depth: 4
    number_sections: false
    theme: "united"
    self_contained: true
    css: "../stylesheet.css"
---

---
date: '`r glue("Document generated: {Sys.Date()}")`'
---

```{r, include = FALSE, echo = FALSE}
if (!require(emo))
  devtools::install_github("hadley/emo")
```

***

Copy, please, these files and directories to your personal directory:

```{bash, eval = FALSE}
cp -r /data/shared/AGE2020/Exercises/E04-microarrays ~/AGE2020/Exercises
```

And switch the R working directory to the current exercise: `setwd("~/AGE2020/Exercises/E04-microarrays")`

***

# Introduction

From [Wikipedia](https://en.wikipedia.org/wiki/DNA_microarray):

> A DNA microarray (also commonly known as DNA chip or biochip) is a collection of microscopic DNA spots attached to a solid surface.
  Scientists use DNA microarrays to measure the expression levels of large numbers of genes simultaneously or to genotype multiple regions of a genome.
  Each DNA spot contains picomoles (10−12 moles) of a specific DNA sequence, known as probes (or reporters or oligos).
  These can be a short section of a gene or other DNA element that are used to hybridize a cDNA or cRNA (also called anti-sense RNA)
  sample (called target) under high-stringency conditions. Probe-target hybridization is usually detected and quantified by
  detection of fluorophore-, silver-, or chemiluminescence-labeled targets to determine relative abundance of nucleic acid sequences in the target.

![A typical protocol for microarray experiment. mRNA is isolated from the sample and reverse transcribed to cDNA which
is labelled by fluorescent dye. The next steps are [hybridization](https://en.wikipedia.org/wiki/DNA_microarray#/media/File:NA_hybrid.svg),
in which cDNA hydridizes with probes on microarray, and the consequent washing, where non-specific sequences are removed, i.e. only strongly paired strands will remain hybridized.
The final wet-lab step is scanning in which the fluorescence intensity of probe spots is measured.
Total strength of the signal, from a spot (feature), depends upon the amount of target sample binding to the probes present on that spot.
Microarrays use relative quantitation in which the intensity of a feature is compared to the intensity of the same feature under
a different condition, and the identity of the feature is known by its position.
Intensities are summarised and analysis of data usually starts with normalization and technical quality control,
and is followed by exploratory analysis and testing for differential expression.
Image adapted from Wikipedia
([1](https://en.wikipedia.org/wiki/DNA_microarray#/media/File:Microarray_exp_horizontal.svg), 
[2](https://commons.wikimedia.org/wiki/File:Microarray_and_sequencing_flow_cell.svg)).](images/microarray.png)

***

# Our experiment

Data come from [experiment](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE57822) where rats were separately treated with 127 compounds dissolved in corn oil.
To control rats only corn oil was given. The goal was to determine the carcinogenity of these compounds.
We will be working only with subset, where are some well known drugs:

- Acetaminophen
- Aspirin
- Diclofenac
- Ibuprofen

For each drug, we have 3 rats (= 3 biological replicates). For controls we have 6 rats.

The used chip is Affymetrix Rat Genome 230 2.0 Array (`Rat230_2`).

***

# Libraries

```{r}
library(oligo)
library(stringr)
library(tidyr)
library(glue)
library(limma)
library(ReportingTools)
library(lattice)
library(sva)
library(dplyr)

source("../age_library.R")
```

***

# Config

```{r}
BASE_DATA_DIR <- "data"
EXPERIMENT_1_DATA_DIR <- glue("{BASE_DATA_DIR}/E-GEOD-57822")
EXPERIMENT_1_SAMPLE_SHEET_FILE <- glue("{EXPERIMENT_1_DATA_DIR}/E-GEOD-57822.sdrf.txt")

# For differential expression analysus report.
REPORT_DIR <- "html_report"
REPORT_N_GENES <- 50
REPORT_LFC_THRESHOLD <- 1
REPORT_P_VALUE_THRESHOLD <- 0.1
REPORT_P_VALUE_ADJUST_METHOD <- "fdr"

# This experiment will be used to demonstrate the batch effect.
EXPERIMENT_2_DATA_DIR <- glue("{BASE_DATA_DIR}/E-GEOD-68065")
```

***

# Reading the data

## Phenotypical data

We need only few columns from the sample sheet. You can explore `r str_replace(EXPERIMENT_1_SAMPLE_SHEET_FILE, ".txt", ".xlsx")` Excel sheet.

```{r}
pheno_data <- readr::read_delim(EXPERIMENT_1_SAMPLE_SHEET_FILE, delim = "\t") %>%
  dplyr::select(1, 48, 41) %>%
  magrittr::set_colnames(c("sample_name", "sample_group", "cel_file"))

pheno_data
```

Let's adjust `pheno_data` a little bit more:

```{r}
pheno_data <- pheno_data %>%
  dplyr::mutate(
    sample_name = str_remove(sample_name, " 1"),
    sample_group = dplyr::if_else(sample_group == "not specified", "control", sample_group),
    sample_group = str_to_lower(sample_group) %>%
      factor() %>%
      relevel("control")
  ) %>%
  as.data.frame() %>%
  magrittr::set_rownames(.$sample_name)

pheno_data
```

## CEL files

Microarray manufacturers use their own format to store intensity data, usually in binary form.
Affymetrix is using [CEL format](http://dept.stat.lsa.umich.edu/~kshedden/Courses/Stat545/Notes/AffxFileFormats/cel.html).

[oligo](https://www.bioconductor.org/packages/release/bioc/html/oligo.html) is a package for analyzing microarrays and supports Affymetrix
([user guide](https://www.bioconductor.org/packages/release/bioc/vignettes/oligo/inst/doc/oug.pdf)).
Raw data are stored in object of class `ExpressionFeatureSet`.

```{r}
cel_files <- glue("{EXPERIMENT_1_DATA_DIR}/{pheno_data$cel_file}")
raw_data <- read.celfiles(cel_files)
```

`raw_data` is object of class `ExpressionFeatureSet`, which is designated to store the raw (non-summarised, non-normalized) data from an experiment.
It is derived from `ExpressionSet` class, of which object we will obtain after data normalization.
`ExpressionSet` is one of the core Bioconductor's classes found in `Biobase` package.
It is used as a container for high-throughput assays and experimental metadata.
You can read its manual [here](https://www.bioconductor.org/packages/release/bioc/vignettes/Biobase/inst/doc/ExpressionSetIntroduction.pdf).

![The basic structure of `ExpressionSet` object.
There are three matrices:
<span style="color: #FF7E7E">intensity (expression) matrix</span> (rows = probes, columns = samples),
<span style="color: #BF7EBF">phenotypical (sample) data</span>, and
<span style="color: #7E7EFF">feature (probe) data</span>, which adds additional information about probes (probe ID, gene ID, symbol, name, etc.).
The <span style="color: #FF7E7E">expression matrix</span> is accesible via `exprs()`.
The <span style="color: #BF7EBF">sample data</span> are accesible via `pData()` (or `phenoData()`, see below).
The <span style="color: #7E7EFF">feature data</span> are accesible via `fData()` (or `featureData()`, see below).
It is important that the latter two matrices are linked to the
<span style="color: #FF7E7E">expression matrix</span> via their rownames:
rownames of the <span style="color: #BF7EBF">phenotypical matrix</span> match **colnames** of the
<span style="color: #FF7E7E">expression matrix</span>, and rownames of the
<span style="color: #7E7EFF">feature matrix</span> match rownames of the
<span style="color: #FF7E7E">expression matrix</span>.
](images/ExpressionSet.png)

There is also other data stored in `ExpressionSet`, such as `experimentData`, but that is not important for now.
You can find it in the manual above or see `?ExpressionSet`.
Alternatively, you can look at methods implemented by a class, in this case the `ExpressionSet`:

```{r}
methods(class = "ExpressionSet")
```

You can see all the stored data in the `raw_data` summary:

```{r}
raw_data
```

Rownames and colnames of `ExpressionSet` can be simply retrieved by the usual functions,
but for better readability is better to use the alternative `Biobase` functions:

```{r}
# or colnames(raw_data)
sampleNames(raw_data)

# or rownames(raw_data)
featureNames(raw_data) %>% head()
```

Rownames are now just row numbers, because we already haven't done the probe summarisation.

Sample and feature data are actually not a basic `data.frame`, but a more complex `AnnotatedDataFrame`.
That is a "special data frame" which consists of two parts:
dataframe describing metadata of samples and dataframe describing this metadata (i.e. its columns).

Let's create such an `AnnotatedDataFrame`.
First we change the sample names to a more neat form:

```{r}
sampleNames(raw_data) <- pheno_data$sample_name
```

Now we create metadata for `AnnotatedDataFrame`:

```{r}
metadata <- data.frame(
  labelName = colnames(pheno_data),
  labelDescription = c("", "Used drug.", ""),
  stringsAsFactors = FALSE
)

metadata
```

And finally the `AnnotatedDataFrame`:

```{r}
(pheno_data <- AnnotatedDataFrame(data = pheno_data, varMetadata = metadata))
```

We also combine `pheno_data` with existing pheno data, as there is some
existing information:

```{r}
phenoData(raw_data)
pData(raw_data) %>% head()
phenoData(raw_data) <- Biobase::combine(phenoData(raw_data), pheno_data)
phenoData(raw_data)
pData(raw_data) %>% head()
```

Look at the difference between `phenoData()` and `pData()`:
the first returns `AnnotatedDataFrame`, while the second `data.frame`.

Accessing the metadata:

```{r}
varMetadata(raw_data)
```

Accessing the expression matrix:

```{r}
exprs(raw_data)[1:5, 1:5]
```

Subsetting work as for normal matrix, but will subset all matrices:

```{r}
raw_data_subset <- raw_data[1:5, 1:5]
raw_data_subset
exprs(raw_data_subset)
pData(raw_data_subset)
fData(raw_data_subset)
```

***

# Technical quality control

Now we will assess whether the experiment was performed correctly at the technical level.

_**Assumption: most of the genes do not change expression across the samples.**_

## Pseudo-image

Pseudo-image  plots  are  used  to  assess  the  spatial  distribution  of  the  data  on  the  chips.
Due to the magnitude of the readings, pseudo-images using data on the original scale often mask spatial features that may be present on the arrays.
This is why we recommend the use of the $log_2$ scale.

```{r}
image(raw_data, which = 12, transfo = log2)
```

You should see something like a random noise.
In the case of a larger area of the same color, there could be some technical problem.

## MA plot

$j, k$ ... arrays

$i$ ... probe

Difference in log-intensities of probe $i$: $M_i=log_2(Y_{i,j}/Y_{i,k})$

Average log-intensity of probe $i$: $A_i=\frac{1}{2}log_2(Y_{i,j}*Y_{i,k})$

Ideally, $M = 0$ should apply for all probes in all samples, due to the assumption that most of the genes do not change expression across the samples..

```{r}
MAplot(raw_data[, 12:14], pairs = TRUE)
```

We can also summarise groups. Defaultly, median is used to summarise probe intensities across the samples.

```{r}
MAplot(raw_data, groups = pData(raw_data)$sample_group, pairs = TRUE)
```

## Boxplots

Ideally, we should see similar boxplots.

```{r}
boxplot(raw_data, nsample = nrow(raw_data))
```

## Probe level model

We will fit PLM using a robust regression which allows to calculate the standard errors of the coefficients.

$log_{2}(Y_{i,j})=\beta_j + \alpha_i + \epsilon_{i,j}$

$Y_{i,j}$ ... background corrected intensity of probe $i$ in sample $j$

$\beta_j$ ... true transcript intensity of sample $j$

$\alpha_i$ ... systematic bias of probe $i$

$\epsilon_{i,j}$ ... random noise of probe $i$ in sample $j$

```{r}
fit_plm <- fitProbeLevelModel(raw_data)
```

Let's see its properties:

```{r}
image(fit_plm, which = 12)
image(fit_plm, which = 12, type = "sign.residuals")
```

Again, as in the case of pseudo-image, you should see something like a random noise.
In the case of a larger area of the same color, there could be some technical problem.

## Relative log expression (RLE)

$y_{i,j}$ ... intensity of gene $i$ in sample $j$

For each gene $i$ we calculate its median intensity across the samples: $median(y_{i,*})$

Then we calculate deviations from this median: $y_{i,j} - median(y_{i,*})$

Ideally, RLE values should be spread around $0$.

```{r}
RLE(fit_plm)
```

## Normalized unscaled standard errors (NUSE)

We can look at the standard error estimates from PLM, for each probe on each array $j$.
They are standardized so that the median standard error across arrays is $1$ for each gene.

$NUSE(\beta_j)=\frac{SE(\beta_j)}{median_j(\beta_j)}$

Ideally, boxplots should be centered around $1$.
Samples having median above $1.1$ and/or higher spread can be considered as low quality ones ([source](http://arrayanalysis.org/main.html)).

```{r}
NUSE(fit_plm)
```

***

# Normalization and probe annotation

As we are working with Affymetrix microarrays, we use the RMA method for normalization:

```{r}
norm_data <- rma(raw_data)
```

Now we only know the probe IDs, but not the gene IDs they correspond to:

```{r}
featureNames(norm_data) %>% head()
```

And so we use [DB package](http://bioconductor.org/packages/release/data/annotation/html/rat2302.db.html) corresponding to our chip to annotate the probes.

All of those annotation packages use SQLite database as a backend and have a common interface called
[`AnnotationDbi`](https://bioconductor.org/packages/release/bioc/html/AnnotationDbi.html).

Some methods in `AnnotationDbi` have names similar to the ones from other packages (like `dplyr::select()`).
So always rather use `AnnotationDbi::` prefix.

Note: because SQLite is a very simple database, it allows only one session (i.e. database file can be opened by only one user at the time).
So it is not possible to use parallel processing when working with DB packages.

```{r}
library(rat2302.db)
rat2302.db
class(rat2302.db)
```

Let's show how to work with `AnnotationDbi` objects.

To show DB columns:

```{r}
AnnotationDbi::columns(rat2302.db)
```

We can show a description of each column:

```{r}
help("UNIGENE")
```

The primary keys of this DB are probe IDs:

```{r}
AnnotationDbi::keys(rat2302.db) %>% head()
```

But there are more key columns to use:

```{r}
AnnotationDbi::keytypes(rat2302.db)
```

And finally, we can select data using our list of keys. We also specify which columns we want to return and what is our keytype:

```{r}
feature_data <- AnnotationDbi::select(
  rat2302.db,
  # Which columns we want to retrieve from DB.
  columns = c("PROBEID", "ENSEMBL", "SYMBOL", "GENENAME", "ENTREZID"),
  # Vector of keys to use. Each key corresponds to its value in different columns.
  keys = featureNames(norm_data),
  # Our key type is ID of probe.
  keytype = "PROBEID"
)

head(feature_data)
```

Some probes are targeting multiple genes:

```{r}
janitor::get_dupes(feature_data, PROBEID) %>% head()
```

How is it possible? Well, I am not sure `r emo::ji("slightly_smiling_face")` But one explanation could be that in the time microarray probes were designed,
they were really targeting a single gene, but some time later, a more precise genome annotation was released.

We have two options: to take the first gene, or to summarise gene information to single row.
In my opinion, the second option is better, as you won't lose any information.
On the other hand, you could have some problems, as, for example, ENSEMBL ID will look like this: `ENSRNOG00000013741, ENSRNOG00000042523`,
and that cannot be used, for example, in joining operations.

For the sake of completeness, we will try both options, but go further in the analysis with the first one.

Let's start with the second option, as we won't use it later.
We group `feature_data` by `PROBEID` and apply function to each of its other four columns.
This function takes separately each of them and joins its values by `, `.

```{r}
feature_data_sum <- feature_data %>%
  dplyr::group_by(PROBEID) %>%
  dplyr::summarise_at(
    vars(ENSEMBL, SYMBOL, GENENAME, ENTREZID),
    ~ str_c(., collapse = ", ")
  )

head(feature_data_sum)
```

And now the first option. We simply drop duplicates with the `dplyr::distinct()` function.
We specify `PROBEID` column to be used for determining uniqueness.

```{r}
feature_data <- dplyr::distinct(feature_data, PROBEID, .keep_all = TRUE)
```

Now we can filter `norm_data` and assign feature data to it:

```{r}
norm_data <- norm_data[feature_data$PROBEID, ]

if (any(feature_data$PROBEID != featureNames(norm_data)))
  stop("Feature data mismatch.")

fData(norm_data) <- feature_data
annotation(norm_data) <- "rat2302.db"

saveRDS(norm_data, file = glue("{EXPERIMENT_1_DATA_DIR}/norm_data.Rds"))
```

Feature data are also stored in `ExpressionSet` as `AnnotatedDataFrame`:

```{r}
featureData(norm_data)
```

But it is not needed to directly assign `AnnotatedDataFrame` - in the step above,
we have assigned just the dataframe with `fData(norm_data) <- feature_data`.

***

# Biological quality control

Now we are ready to look at some biology.
It is time to reuse your functions `r emo::ji("slightly_smiling_face")`

```{r}
groups <- pData(norm_data)$sample_group
names(groups) <- sampleNames(norm_data)
plot_hc(
  exprs(norm_data), color_by = groups,
  color_by_lab = "Sample Group"
)

plot_pca_ggplot2(exprs(norm_data), pData(norm_data), color_by = "sample_group")$plot
```

For heatmap, 1000 probes are enough. Don't use a big number of probes, or your browser will freeze `r emo::ji("wink")`

```{r}
plot_pheatmap(exprs(norm_data)[1:1000, ], pData(norm_data), column_color_by = "sample_group", main = "Affymetrix", show_rownames = FALSE)
```

***

# Differential expression analysis (DEA)

DEA tells us which genes are differentially expressed between conditions.
As in the qPCR, this is again a relative measure: for example, we can say that a gene is third times
more expressed in treatment group, relative to control group.

In the qPCR, we were using the simple t-test to test for the differential expression.
We could use that, because qPCR data are small enough to do so.
On the other hand, data from the high-throughput methods, such as microarrays and RNA-seq, are
much bigger and one of their properties is that the number of samples is always **much** smaller than the number of features.
For that reason, we have to do DEA using a more complex statistical method: linear models.

## `limma`

For DEA we will be using [limma](https://bioconductor.org/packages/release/bioc/html/limma.html).
It is the most known package (Bioconductor top 12. package, `r format(Sys.time(), "%Y-%d-%m")`)
for DEA on microarray data and was probably the first which uses linear models for that.

We will fit a linear model and improve estimates of standard errors using empirical Bayes approach.
Because we have 5 sample groups (the first one being control), our model looks like this:

$Y_i = \beta_0 + \beta_1 x_{i,1} + \dotso + \beta_5 x_{i,5} + \epsilon_i, i = 1 \dotso N$

This can be rewritten to matrix form as:

$$
  \begin{pmatrix}
    Y_1 \\ Y_2 \\
    \vdots \\
    Y_N\end{pmatrix} =
  \begin{pmatrix}
    1 & x_{1,1} & \dotso & x_{1,5} \\
    \vdots & \vdots & \ddots & \vdots \\
    1 & x_{N,1} & \dotso & x_{N,5}
  \end{pmatrix}
  \begin{pmatrix}
    \beta_0 \\
    \beta_1 \\
    \vdots \\
    \beta_5
  \end{pmatrix} +
  \begin{pmatrix}
    \epsilon_1 \\
    \epsilon_2 \\
    \vdots \\
    \epsilon_N
  \end{pmatrix}
$$

It can be simplified to:

$\mathbf{Y} = \mathbf{X \beta} + \mathbf{\epsilon}$

The matrix $\mathbf{X}$ is the design matrix. Let's make it:

```{r}
group <- pData(norm_data)$sample_group %>% factor() %>% relevel("control")
dea_model <- model.matrix(~ group)
colnames(dea_model)[1] <- "Intercept"
dea_model
```

So for example, how is the first sample of *diclofenac* group expressed?

$Y_{diclofenac,1} = \beta_0*1 + \beta_1*0 + \beta_2*0 + \beta_3*1 + \beta_4*0 + \epsilon = \beta_0 + \beta_3 + \epsilon$

Intercept is our $\beta_0$ (i.e. *control* = the reference group).
In factor variable, the first level is automatically taken as the reference level.

Some models could be quite complicated. You can read more about model design [here](http://genomicsclass.github.io/book/) (Chapter 5).

Now we can fit the model and use the empirical Bayes shrinkage:

```{r}
fit <- lmFit(norm_data, dea_model) %>% eBayes()
colnames(fit)
```

Using the `topTable()` function, we can look at differential expression statistics for each gene in a specific group
(in this case, *acetaminophen* vs *control*). We call those comparisons **contrasts**.

```{r}
topTable(fit, coef = "groupacetaminophen")
```

Let's inspect the `topTable` columns:

- `logFC`: $log_2(\text{fold-change})$ relative to reference group (in this case `control`).
- `AveExpr`: average expression in group (in this case `acetaminophen`).
- `t`: t-statistic.
- `P.Value`: p-value of the differential expression test.
- `adj.P.Val`: p-value adjusted for multiple testing issue.
- `B`: log-odds that the gene is differentially expressed.

We didn't specify any parameter to `topTable`, but if you look at its help page,
you can see that it is possible to set p-value adjust method and also to set thresholds for maximum
adjusted p-value, as well as for minimal absolute value of LFC.
By default, it won't filter anything and return DEA statistics for first 10 genes by their `B` value.

## DEA between other sample groups

But what if we want to compare two non-control groups?
Well, we need to specify our own contrasts (in this case, *aspirin* vs *ibuprofen* and *acetaminophen* vs *diclofenac*):

```{r}
contrasts <- makeContrasts(
  aspirin_vs_ibuprofen = groupaspirin - groupibuprofen,
  acetaminophen_vs_diclofenac = groupacetaminophen - groupdiclofenac,
  levels = dea_model
)

contrasts
```

Behind it is just $(aspirin - control) - (ibuprofen - control) = aspirin - ibuprofen$, so comparison to control (reference) group simply cancels out.
Now we can refit our model using our defined contrasts:

```{r}
fit_c <- contrasts.fit(fit, contrasts = contrasts) %>% eBayes()
colnames(fit_c)
topTable(fit_c, coef = "aspirin_vs_ibuprofen")
```

## ReportingTools

[ReportingTools](https://bioconductor.org/packages/release/bioc/html/ReportingTools.html) is a great package for HTML reporting of DEA.
Basically, it outputs a table from the `topTable()`, but in an interactive form.
It is working with many DE analysis packages, as you will see in RNA-seq.

We will iterate over all treatment sample groups and for each we create the HTML report.
Because `ReportingTools` are throwing error when no probe is meeting a filtering criteria,
we have to put the code inside the `tryCatch()` function.
The following code could be a bit complicated, so don't hesitate to ask me.

```{r, results = "asis"}
output_groups <- levels(groups)
output_groups <- output_groups[output_groups != "control"]

rep_theme <- reporting.theme()
lattice.options(default.theme = rep_theme)

for (group in output_groups) {
  cat(glue("\nDEA for group '{group}': ", .trim = FALSE))
  
  tryCatch(
    {
      # This is an object (and a file) to which we are publishing.
      de_report <- HTMLReport(
        shortName = group,
        title = glue("{group} vs. control"),
        reportDirectory = REPORT_DIR
      )
      
      # We can control what to take from topTable.
      # Here, we want top REPORT_N_GENES DE genes (ranked by p-value adjusted by FDR) with minimal LFC of 1 (i.e. two-fold up or down).
      # We won't be much conservative and set adjusted p-value cutoff to 0.1. That means we expect 10% of our DE genes to be false positives.
      publish(
        fit,
        de_report,
        eSet = norm_data,
        factor = groups,
        coef = glue("group{group}"),
        n = REPORT_N_GENES,
        lfc = REPORT_LFC_THRESHOLD,
        pvalueCutoff = REPORT_P_VALUE_THRESHOLD,
        adjust.method = REPORT_P_VALUE_ADJUST_METHOD
      )
      
      finish(de_report)
      
      cat(glue("<a href='{DE_REPORT_DIR}/{group}.html' target='_blank'>report</a>"))
    },
    
    error = function(e) {
      cat(e$message)
  })
}
```

However, for a HTML report to be loaded correctly, you must open it from normal filesystem, not RStudio file browser.
So copy the `html_report` directory to your local computer or access the HTML files with `sshfs`.

***

# Boxplots

We can use our boxplot functions also for microarray data.
In case of `plot_boxplot_ggplot2()`, we need to convert `ExpressionSet` to the long format.
And this is a big drawback, because such long dataframe will be **huge**.
So don't convert a full dataset at once, unless you have a lot of RAM.

We convert to the long format only the subset of first 10 probes:

```{r}
data_long <- exprs(norm_data)[1:10, ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("PROBEID") %>%
  tidyr::pivot_longer(-PROBEID, names_to = "sample_name", values_to = "E") %>%
  dplyr::left_join(fData(norm_data), by = "PROBEID") %>%
  dplyr::left_join(pData(norm_data), by = "sample_name")

data_long
```

Let's make a boxplot:

```{r}
plot_boxplot_ggplot2(
  data_long,
  x = "sample_group",
  y = "E",
  feature_col = "SYMBOL",
  color_by = "sample_group",
  main = "Affymetrix",
  x_lab = "Sample_Group",
  y_lab = "log2(expression intensity)",
  do_t_test = FALSE
) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

As for the qPCR data, it would be nice to see LFC between the treatment groups and the `control` group.
Again, this is doable by subtracting the mean expression intensity of the `control` group.

***

# Batch effect

Batch effects are technical sources of variation that have been added to the samples during handling or,
in our case, by different technical preparation of the experiments.
They have to be removed, otherwise they act as confounding elements which can be wrongly interpreted as biological variation.

We will join two experiments. The one we have been working with and the [second one](https://www.ebi.ac.uk/arrayexpress/experiments/E-GEOD-68065/)
in which rats were treated with acetaminophen.

The preprocessing is quite similar to what we have already done.

```{r}
SAMPLE_SHEET_2_FILE <- glue("{EXPERIMENT_2_DATA_DIR}/E-GEOD-68065.sdrf.txt")

pheno_data_2 <- readr::read_delim(SAMPLE_SHEET_2_FILE, delim = "\t") %>%
  dplyr::select(1, 4, 5, 38) %>%
  magrittr::set_colnames(c("sample_name", "sample_group", "rat_status", "cel_file"))

head(pheno_data_2)
```

```{r}
pheno_data_2 <- pheno_data_2 %>%
  dplyr::filter(rat_status == "susceptible") %>%
  dplyr::mutate(
    sample_name = str_replace(sample_name, " 1", ""),
    sample_group = dplyr::if_else(str_detect(sample_group, "predose"), "control", "acetaminophen"),
    batch = 2,
    rat_status = NULL
  )

pheno_data_2
```

We modify the sample sheet from first experiment to contain only

```{r}
pheno_data_batch <- pData(norm_data) %>%
  dplyr::filter(sample_group %in% c("control", "acetaminophen")) %>%
  dplyr::mutate(batch = 1) %>%
  dplyr::bind_rows(pheno_data_2) %>%
  dplyr::mutate(
    index = NULL,
    sample_group = factor(sample_group) %>% relevel("control"),
    batch = factor(batch),
    cel_file = dplyr::if_else(batch == 1, glue("{EXPERIMENT_1_DATA_DIR}/{cel_file}"), glue("{EXPERIMENT_2_DATA_DIR}/{cel_file}"))
  ) %>%
  as.data.frame() %>%
  magrittr::set_rownames(.$sample_name)

pheno_data_batch
```

```{r}
raw_data_batch <- read.celfiles(pheno_data_batch$cel_file)

sampleNames(raw_data_batch) <- pheno_data_batch$sample_name
metadata <- data.frame(
  labelName = colnames(pheno_data_batch),
  labelDescription = c(""),
  stringsAsFactors = FALSE
)
pheno_data_batch <- AnnotatedDataFrame(data = pheno_data_batch, varMetadata = metadata)
phenoData(raw_data_batch) <- Biobase::combine(phenoData(raw_data_batch), pheno_data_batch)
```

```{r}
MAplot(raw_data_batch, groups = pData(raw_data_batch)$sample_group, pairs = TRUE)

boxplot_colors <- dplyr::if_else(pData(raw_data_batch)$batch == 1, "red", "blue")
boxplot(raw_data_batch, nsample = nrow(raw_data_batch), col = boxplot_colors)
legend("topright", legend = c("1", "2"), title = "Batch", col = c("red", "blue"), pch = 20)
```

```{r}
fit_plm_batch <- fitProbeLevelModel(raw_data_batch)
RLE(fit_plm_batch)
NUSE(fit_plm_batch)
```

Hmm, it looks suspicious, doesn't it? `r emo::ji("thinking")`

```{r}
norm_data_batch <- rma(raw_data_batch)
```

Here, at the biological level, you can see that the samples are obviously divided according to batches.

```{r}
groups <- pData(norm_data_batch)$sample_group
batches <- pData(norm_data_batch)$batch
names(groups) <- sampleNames(norm_data_batch)
names(batches) <- sampleNames(norm_data_batch)

plot_hc(exprs(norm_data_batch), color_by = groups, color_by_lab = "Sample Group")
plot_hc(exprs(norm_data_batch), color_by = batches, color_by_lab = "Batch")
plot_pca_ggplot2(exprs(norm_data_batch), pData(norm_data_batch), color_by = "sample_group", shape_by = "batch")$plot
plot_pheatmap(
  exprs(norm_data_batch)[1:1000, ],
  pData(norm_data_batch),
  column_color_by = c("sample_group", "batch"),
  main = "Affymetrix",
  show_rownames = FALSE
)
```

Let's correct the batch effect with the `ComBat()` function from `sva` package.
We will specify how are samples divided to batches and what is our experimental design.

```{r}
norm_data_batch_combat <- norm_data_batch
exprs(norm_data_batch_combat) <- ComBat(exprs(norm_data_batch_combat), batch = batches, mod = model.matrix(~ groups))
```

Let's look again at biological exploratory analysis. The batch effect seems to be corrected `r emo::ji("slightly_smiling_face")`

```{r}
plot_hc(exprs(norm_data_batch_combat), color_by = groups, color_by_lab = "Sample Group")
plot_hc(exprs(norm_data_batch_combat), color_by = batches, color_by_lab = "Batch")
plot_pca_ggplot2(exprs(norm_data_batch_combat), pData(norm_data_batch_combat), color_by = "sample_group", shape_by = "batch")$plot
plot_pheatmap(
  exprs(norm_data_batch_combat)[1:1000, ],
  pData(norm_data_batch_combat),
  column_color_by = c("sample_group", "batch"),
  main = "Affymetrix",
  show_rownames = FALSE
)
```

# Cleanup

```{r, warning = TRUE, message = TRUE}
save.image("microarrays.RData")

warnings()
traceback()
sessionInfo()
```

# HTML rendering

This chunk is not evaluated (`eval = FALSE`). Otherwise you will probably end up in recursive hell `r emo::ji("exploding_head")`

```{r, eval = FALSE, message = FALSE, warning = FALSE}
library(rmarkdown)
library(knitr)
library(glue)

# You can set global chunk options. Options set in individual chunks will override this.
opts_chunk$set(warning = FALSE, message = FALSE)
render("microarrays.Rmd", output_file = "microarrays.html")
```
